Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-09 mr-review.Rmd
added text

Methods for and uses of MR have appeared at a rapid and accelerating pace since the 
appearance of the early GWAS (Figure \@ref(fig:article-counts)). Inverse variance 
weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted 
median estimation [@bowden_consistent_2016], Bayesian weighted MR 
[@zhao_bayesian_2019], robust adjusted profile score [@zhao_statistical_2020], MRMix 
[@qi_mendelian_2019], CAUSE [@morrison_mendelian_2020], and MRAID 
[@yuan_likelihood_2021] are all recent methods developments. These methods differ in 
their approaches to two challenges:

@@

Recent developments have enabled execution of two-sample MR studies with only 
GWAS summary statistics [@burgess2015using]. These methods are especially pertinent now that there is 
an abundance of freely available GWAS summary data [@buniello2019nhgri]. The 
ability to use published has led to a rapid increase in the number of available
two-sample MR methods for GWAS summary statistics.

@@

The outcome variable can be binary, a count, or a continuous variable. For brevity and simplicity, 
we focus on continuous outcomes below.

@@

If the first assumption, about the association between the genetic marker and the exposure, holds, 
but this association is weak, then an amplification of biases, such as those due to violations of 
assumptions 2 and 3, may result [@vanderweele_methodological_2014]. @burgess_avoiding_2011, in analyses 
of simulated and real data, demonstrated that bias in causal estimates increases with decreases in 
the strength of association between the SNP and exposure.

The second assumption, often termed the exclusion restriction assumption, can also be stated as 
the need for the genetic marker to affect the outcome only through the exposure. In a 
causal diagram, this means that the only path from the marker to the outcome is that 
containing the exposure.

@@

When the three MR assumptions hold, one may use instrumental variable statistical methods, 
with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the 
causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008].

@@

A two-stage statistical model involves two regression models [@burgess_review_2017]. For a continuous 
outcome, one may choose to perform "two-stage least squares", which involves two linear regressions. 
First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)). The resulting 
fitted values for the exposure variable are the independent variable in the second linear regression, 
where the outcome variable is the dependent variable. The causal effect estimate, then, is the 
coefficient $\hat \beta_1$ obtained from the second regression analysis (Equation \@ref(eq:lm2)).

@@

# Mendelian Randomization with multiple instruments

Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR 
methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011;@bowden_mendelian_2015].

Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
in MR studies. The solution to this task differs among published methods. Early methods chose a 
set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
SNPs in LD with each other. In this setting, discarding some of the causal SNPs may only capture a small
proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
this is the finding that including multiple correlated SNPs in transcriptome-wide association
studies improves power compared to using independent SNPs only [@liu2021multi;@gusev_integrative_2016;@yuan_testing_2020;@zeng_non-parametric_2017].

@@

In the case of continuous exposures and outcomes, researchers have taken several approaches to extend the methods and models for 
single-SNP MR to accommodate multiple SNPs. For two stage analyses, one may perform 
a sequence of two linear regressions, as in the single-SNP MR models (Equations \@ref(eq:lm3) and \@ref(eq:lm4)). 
For an analysis
with $S$ SNP instruments, one performs two stage analysis by first regressing the 
exposure $n$ vector $X$ on the $n$ by $S$ matrix of SNP instruments. Taking the fitted exposure values 
from this regression, the second stage involves regressing the outcome variable $Y$ on 
the fitted values, $\hat X$.

\begin{equation}
  X  = \alpha Z + \epsilon
  (\#eq:lm3)
\end{equation}

@@

\begin{equation}
  Y = \beta \hat X + \tau
  (\#eq:lm4)
\end{equation}

@@

Horizontal pleiotropy refers to the presence of multiple pathways from a genetic 
variant to the outcome. One of these multiple pathways may contain the exposure variable.
The fact that there are multiple pathways from the genetic variant to the outcome violates
standard MR assumptions. @verbanck_detection_2018 documented widespread horizontal pleiotropy 
in GWAS. This finding motivated the development of MR-PRESSO and inspired 
further research into methods to model horizontal pleiotropy [@verbanck_detection_2018].

Researchers have recognized two types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
SNP-exposure relationships. The second type of horizontal pleiotropy manifests when 
in the presence of unobserved exposure-outcome confounding. It induces correlation between 
horizontal pleiotropic effects and SNP-exposure effects. Both types of horizontal pleiotropy 
violate standard MR modeling assumptions and can bias causal effect estimates and can increase false 
discoveries [@yuan_likelihood_2021].
Early MR analyses avoided confounding from horizontal pleiotropy by discarding 
instrumental SNPs that might be associated with the outcome 
[@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
More recent methods have attempted to model horizontal pleiotropy 
[@morrison_mendelian_2020;@yuan_testing_2020;@yuan_likelihood_2021]. CAUSE [@morrison_mendelian_2020] and 
MRMix [@qi_mendelian_2019] both use a mixture of normal distributions to control for both 
types of horizontal pleiotropy. Modeling both types of horizontal pleiotropy is particularly 
challenging because the MR model likelihood often involves an integral that can't be solved analytically. 
Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.
MRMix uses ...
CAUSE compares the out-of-sample prediction accuracies for two competing models that differ in inclusion of 
the causal effect parameter. Specifically, CAUSE calculates the expected log pointwise posterior density 
between the two models. 
MRAID models horizontal pleiotropy by ...
@yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is 
biased. CAUSE, they found, yields overly conservative p-values.

>>>>

Methods for and uses of MR have appeared at a rapid and accelerating pace since the 
publication of the early GWAS (Figure \@ref(fig:article-counts)). Inverse variance 
weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted 
median estimation [@bowden_consistent_2016], Bayesian weighted MR 
[@zhao_bayesian_2019], robust adjusted profile score [@zhao_statistical_2020], MRMix 
[@qi_mendelian_2019], CAUSE [@morrison_mendelian_2020], and MRAID 
[@yuan_likelihood_2021] are all recent methods developments. These methods differ in 
their approaches to two challenges:

@@

Horizontal pleiotropy occurs when a SNP instrument affects the outcome through at least one pathway that bypasses the exposure variable [@verbanck_detection_2018]. It constitutes a violation of 
the standard MR assumptions and can lead to biased 
causal effect estimates and diminished statistical power. We discuss approaches to 
modeling horizontal pleiotropy below.

Recent methods developments have enabled execution of two-sample MR studies with only 
GWAS summary statistics [@burgess2015using]. These strategies are especially pertinent now that there is 
an abundance of freely available GWAS summary data [@buniello2019nhgri]. The 
ability to use published data has contributed to a rapid increase in the number of
two-sample MR methods for GWAS summary statistics.

@@

The outcome variable can be binary, a count, a time to event, or a continuous variable. For brevity and simplicity, 
we focus on continuous outcomes before considering other classes of outcome variables in a later section.

@@

If the first assumption, about the association between the genetic marker and the exposure, holds, 
but this association is weak, then an amplification of biases, such as those due to violations of 
assumptions 2 and 3, may result [@vanderweele_methodological_2014]. @burgess_avoiding_2011, in analyses 
of simulated and real data, demonstrated that bias in causal estimates increases with decreases in 
the strength of association between the SNP and exposure. We later discuss, in the context of 
methods for multiple SNP instruments, how this finding, coupled with the recognition of 
the polygenicity of many complex traits, has motivated methods to work with large numbers of weak SNP instruments [@ye_debiased_2021;@chao_consistent_2005].

The second MR assumption, often termed the exclusion restriction assumption, can also be stated as 
the need for the genetic marker to affect the outcome only through the exposure. In a 
causal diagram, this means that the only path from the marker to the outcome is that 
containing the exposure.

@@

When the three MR assumptions hold, one may use instrumental variable statistical methods, 
with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the 
causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008]. It's important 
to recognize that assumptions 2 and 3, which require absence of two types of confounding, 
can't be verified in observational studies. For this reason, researchers have developed 
sensitivity analysis to consider the robustness of their inferences to possible violations 
of assumptions 2 and 3.

@@

A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
outcome, one may choose to perform "two-stage least squares", which involves two linear regressions. 
First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)). The resulting 
fitted values for the exposure variable are the independent variable in the second linear regression, 
where the outcome variable is the dependent variable. The causal effect estimate, then, is the 
coefficient $\hat \beta_1$ obtained from the second regression analysis (Equation \@ref(eq:lm2)).

@@

# Mendelian Randomization with multiple SNP instruments

Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR 
methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011;@bowden_mendelian_2015]. 
Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
in MR studies. The solution to this task differs among published methods. Early methods chose a 
set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
SNPs in LD with each other. In this setting, discarding some of the causal SNPs may only capture a small
proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
this is the finding that including multiple correlated SNPs in transcriptome-wide association
studies improves power compared to using independent SNPs only [@liu2021multi;@gusev_integrative_2016;@yuan_testing_2020;@zeng_non-parametric_2017].

@@

In the case of continuous exposures and outcomes, researchers have taken several approaches to extend the methods and models for 
single-SNP MR to accommodate multiple SNPs. For two stage analyses, one may perform 
a sequence of two linear regressions, as in the single-SNP MR models (Equations \@ref(eq:lm3) and \@ref(eq:lm4)). 
For an analysis
with $S$ SNP instruments, one performs two stage analysis by first regressing the 
exposure $n$ vector $X$ on the $n$ by $S + 1$ matrix of SNP instruments (and an intercept). Taking the fitted exposure values 
from this regression, the second stage involves regressing the outcome variable $Y$ on 
the fitted values, $\hat X$.

\begin{equation}
  X  = Z\alpha + \epsilon
  (\#eq:lm3)
\end{equation}

@@

\begin{equation}
  Y = \hat X\beta + \tau
  (\#eq:lm4)
\end{equation}

@@

Pleiotropy, where a single genetic variant affects multiple traits, has a long history of study in genetics and complex traits [@solovieff_pleiotropy_2013;@visscher_plethora_2016;@stearns_one_2010]. Horizontal pleiotropy refers to the presence of multiple causal pathways from a genetic 
variant to the outcome. 
The fact that there are pathways that exclude the exposure and connect the SNP to the outcome violates
standard MR assumptions. @verbanck_detection_2018 documented widespread horizontal pleiotropy 
in GWAS. This finding motivated the development of MR-PRESSO and inspired 
further research into methods to model horizontal pleiotropy [@verbanck_detection_2018].

Researchers have recognized two types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
SNP-exposure relationships. The second type of horizontal pleiotropy manifests when 
in the presence of unobserved exposure-outcome confounding. It induces correlation between 
horizontal pleiotropic effects and SNP-exposure effects. Both types of horizontal pleiotropy 
violate standard MR modeling assumptions and can bias causal effect estimates and can increase false 
discoveries [@yuan_likelihood_2021].
Early MR analyses avoided confounding from horizontal pleiotropy by discarding 
instrumental SNPs that might be associated with the outcome 
[@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
More recent methods have attempted to model horizontal pleiotropy 
[@morrison_mendelian_2020;@yuan_testing_2020;@yuan_likelihood_2021]. CAUSE [@morrison_mendelian_2020] and 
MRMix [@qi_mendelian_2019] both use a mixture of normal distributions to control for both 
types of horizontal pleiotropy. Modeling both types of horizontal pleiotropy is particularly 
challenging because the MR model likelihood often involves an integral that can't be solved analytically. 
Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.
MRMix uses ...
CAUSE compares the out-of-sample prediction accuracies for two competing models that differ in inclusion of 
the causal effect parameter [@morrison_mendelian_2020]. Specifically, CAUSE calculates the expected log pointwise posterior density 
between the two models. 
MRAID models horizontal pleiotropy by ...
@yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is 
biased. CAUSE, they found, yields overly conservative p-values [@yuan_likelihood_2021].
